---
title: "ps5-markdown"
author: "Zareen Tasnim"
date: "2023-02-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("gapminder.csv.bz2")
```

## Load and Check Data

2. (2pt) Load data. How many rows/columns do we have?
```{r, message=FALSE}
nrow(gapminder)
ncol(gapminder)
```
There are 13055 rows and 25 columns.

3. (2pt) Print a small sample of data. Does it look OK?
```{r, message=FALSE}
gapminder %>% 
  sample_n(10)
```
The data does look correct.


## Descriptive Statistics

1. (3pt) How many countries are there in the dataset? Analyze all three: iso3, iso2 and name.
```{r, message=FALSE}
gapminder %>% 
  select(name, iso3, iso2) %>% 
  summarize(count_name = n_distinct(name), count_iso3 = n_distinct(iso3), count_iso2 = n_distinct(iso2))
```

There are 250 countries in this data set.

2. If you did this correctly, you saw that there are more names than iso-2 codes, 
and there are even more iso3 -codes. What is going on? Can you find it out?
(a) (5pt) Find how many names are there for each iso-2 code. Are there any iso-2 codes that 
correspond to more than one name? What are these countries?
```{r, message=FALSE}
gapminder %>% 
  group_by(iso2) %>% 
  summarize(num = n_distinct(name, na.rm = TRUE)) %>% 
  arrange(desc(num))
```
There is 1 name per country.

(b) (5pt) Now repeat the same for name and iso3-code. Are there country names that have
more than one iso3-code? What are these countries?
Hint: two of these entities are CHANISL and NLD CURACAO.
```{r, message=FALSE}
gapminder %>% 
  group_by(name) %>% 
  summarize(num = n_distinct(iso3, na.rm = TRUE)) %>% 
  arrange(desc(num))
```
Afghanistan has more than one name. 

3. (2pt) What is the minimum and maximum year in these data?
```{r, message=FALSE}
summary(gapminder$time)
```
The minimum year is 1960 and the maximum year is 2019.


## CO2 Emissions

Next, let’s analyze CO2 emissions.
1. (2pt) How many missing co2 emissions are there for each year? Analyze both missing CO2
and co2_PC. Which years have most missing data?
```{r, message=FALSE}
gapminder %>% 
  group_by(time) %>% 
  select(co2) %>% 
  filter(is.na(co2)) %>% 
  count()

gapminder %>% 
  group_by(time) %>% 
  select(co2_PC) %>% 
  filter(is.na(co2_PC)) %>% 
  count()
```
2017-2019 have the most missing data.

2. (5pt) Make a plot of total CO2 emissions over time for the U.S, China, and India. 
Add a few more countries of your choice. Explain what do you see.
```{r}
countries_overtime <- gapminder %>% 
  filter(name %in% c("United States of America", "China", "India", "Bangladesh", "Japan", "Australia"))

ggplot(countries_overtime, aes(time, co2, color = name)) +
  geom_line(size = 1.2) +
  labs(x = "Time", 
       y = "Total CO2 Emissions", 
       title = "Total CO2 Emissions Over Time",
       color = "Country") 
```
I notice that China has has the most exponential growth in CO2 emissions, the US has always been high steadily but increased little by little. I also saw that some countries don't have much data or just don't contribute much emissions like Bangladesh and Australia.

3. (5pt) Now let’s analyze the CO2 emissions per capita (co2_PC ). 
Make a similar plot of the same countries. What does this figure suggest?
```{r, message=FALSE}
ggplot(countries_overtime, aes(time, co2_PC, color = name)) +
  geom_line(size = 1.2) +
  labs(x = "Time", 
       y = "CO2 Emissions Per Capita", 
       title = "Total CO2 Emissions per Capita",
       color = "Country") 
```
This plot suggests that The higher your capita, the more emissions you have. It's true for the US and Australia clearly. Bangladesh and India are very low contributes and have low per capita.

4. (6pt) Compute average CO2 emissions per capita across the continents. Comment what do you see.
Note: just compute averages over countries and ignore the fact that countries are of different size.
Hint: Americas 2016 should be 4.80.
```{r, message=FALSE}
gapminder %>% 
  filter(time == '2016') %>% 
  group_by(region) %>% 
  summarize(co2_pc_avg = mean(co2_PC, na.rm = TRUE))
```
I notice that a lot of countries are unaccounted for. Asia and Europe seem to be the largest contributers while Africa is least.

5. (7pt) Make a barplot where you show the previous results–average CO2 emissions per capita 
across continents in 1960 and 2016.
```{r, message=FALSE}
avg <- gapminder %>% 
  filter(time == '2016' & time == '1960') %>% 
  group_by(region) %>% 
  summarize(co2_pc_avg = mean(co2_PC, na.rm = TRUE))

  ggplot(avg, aes(region, co2_pc_avg, fill = factor(time))) +
  geom_bar(stat = "identify", position = "dodge") +
    labs(title = "Average CO2 Emissions per Capita Across Continents",
         x = "Continent",
         y = "Average CO2 Emissions per Capita",
         fill = "time")
```


6. Which countries are the three largest, and three smallest CO2 emitters 
(in terms of CO2 per capita) in 2019 for each continent? (Assume region is continent).

```{r, message=FALSE}
gapminder %>%
  filter(!is.na(co2_PC)) %>%
  filter(time == "2016") %>%
  group_by(region, name) %>%
  summarize(avg_co2_pc = mean(co2_PC), .groups = "drop") %>%
  arrange(region, avg_co2_pc) %>%
  group_by(region) %>%
  head(3) %>%
  select(region, name, avg_co2_pc)
```



## GDP per Capita

1. (8pt) Make a scatterplot of GDP per capita versus life expectancy by country, using data for
1960. Make the point size dependent on the country size, and color those according to the
continent. Feel free to adjust the plot in other ways to make it better.
Comment what do you see there.
```{r,message=FALSE}
gapminder %>% 
  
```


2. (4pt) Make a similar plot, but this time use 2019 data only.
```{r, message=FALSE}

```


3. (6pt) Compare these two plots and comment what do you see. How has world developed
through the last 60 years?
```{r, message=FALSE}

```


4. (6pt) Compute the average life expectancy for each continent in 1960 and 2019. Do the results
fit with what do you see on the figures?
Note: here as average I mean just average over countries, ignore the fact that countries are of
different size.
```{r, message=FALSE}

```


5. (8pt) Compute the average LE growth from 1960-2019 across the continents. Show the results
in the order of growth. Explain what do you see.
Hint: these data (data in long form) is not the simplest to compute growth. But you may
want to check out the lag() function. And do not forget to group data by continent when
using lag(), otherwise your results will be messed up! See https://faculty.washington.
edu/otoomet/info201-book/dplyr.html#dplyr-helpers-compute.
```{r, message=FALSE}

```


6. (6pt) Show the histogram of GDP per capita for years of 1960 and 2019. Try to put both
histograms on the same graph, see how well you can do it!
```{r, message=FALSE}

```


7. (6pt) What was the ranking of US in terms of life expectancy in 1960 and in 2019? (When
counting from top.)
Hint: check out the function rank()!
Hint2: 17 for 1960.
```{r, message=FALSE}
gapminder %>%
    select(name, time, lifeExpectancy) %>% 
    filter(name != "", !is.na(time), !is.na(lifeExpectancy)) %>%
    filter(time == "1960") %>% 
    mutate(us_rank = rank(desc(lifeExpectancy))) %>% 
    filter(name == "United States of America")
```


8. (6pt) If you did this correctly, then you noticed that US ranking has been falling quite a
bit. But we also have more countries in 2019–what about the relative rank divided by the
corresponding number of countries that have LE data in the corresponding year?
Hint: 0.0904 for 1960
```{r, message=FALSE}

```



## Finally, how many hours did you spend?

I spent about 23 hours on this.













